SPARQL on the Open, Decentralised Web

### 2016-06-23 dissertation/introduction.tex
Turtle example

>>>>

The use of IRIs makes it possible to view RDF statements as a directed
graph, where the subject IRI is a vertex, the predicate is an edge to
an object vertex\footnote{In general it is more precisely described as a
  partially labelled directed pseudograph with unique node labels}.

For historical reasons, the first syntax that was standardised for RDF
was an syntax based on the tree-structured Extensible Markup Language
(XML). The simplest standardised syntax is known as N-Triples, where
each statement is written on a single line, where each IRI is written
out in full. The basic building blocks of N-Triples has been reused to
create a compact and natural text form syntax, known as Turtle. Turtle
has gained traction, and is well established as a common syntax. A
simple graph expressed in Turtle could be:

\begin{verbatim}
@prefix foaf: <http://xmlns.com/foaf/0.1/>.
@prefix dct:  <http://purl.org/dc/terms/>.

<http://www.kjetil.kjernsmo.net/foaf#me> a foaf:Person ;
    foaf:name "Kjetil Kjernsmo" ;
    foaf:publications <http://folk.uio.no/kjekje/2013/iswc.pdf> .
<http://folk.uio.no/kjekje/2013/iswc.pdf> a foaf:Document ;
    dct:title "Introducing Statistical Design of Experiments to SPARQL Endpoint Evaluation" ; 
    dct:creator <http://www.kjetil.kjernsmo.net/foaf#me>, 
                <http://data.semanticweb.org/person/john-s-tyssedal> .
\end{verbatim}

The example opens with prefix declarations, to allow abbreviating
IRIs, e.g. the predicate \rdfterm{foaf:name} is an abbreviation of the
IRI \rdfterm{http://xmlns.com/foaf/0.1/name}. The IRI
\rdfterm{http://www.kjetil.kjernsmo.net/foaf#me} is a name for this
author. The \rdfterm{a} predicate is standards-defined abbreviation
for \rdfterm{http://www.w3.org/1999/02/22-rdf-syntax-ns#type}, which
is used for declaring the subject an instance of a class, in this case
\rdfterm{foaf:Person}. Then, the \rdfterm{foaf:publications} predicate
is used to link to an author-published version of one of the
publications of this dissertation, and further information is then
provided about this publication, including a link to the co-author. As
we see, semicolons are used to separate statements that share a
subject, commas separate statements with the same subject and
predicate, while periods terminate a statement.

Also, note that all IRIs in this example are \emph{dereferenceable},
meaning that if an HTTP client is used, a representation of the
resource can be retrieved across the Internet. This, along with that
it is expressed as RDF, makes the above so-called Linked Data.
